---
title: "Visualising Coronavirus"
output:
  html_document:
    df_print: paged
    toc: true
---

# Libraries and Data

```{r}
library(pacman)
pacman::p_load(
  tidyquant,
  readr,
  ggplot2,
  dplyr,
  knitr,
  ggthemes,
  lubridate
)

# data available at
# https://www.ecdc.europa.eu/en/publications-data/download-todays-data-geographic-distribution-covid-19-cases-worldwide
data <- read.csv("https://opendata.ecdc.europa.eu/covid19/casedistribution/csv", na.strings = "", fileEncoding = "UTF-8-BOM")
```

# Covid Library

[Documentation](https://github.com/RamiKrispin/coronavirus)

Another useful way to get data and daily updates is with the covid library which can be installed as follows:


```{r,eval=FALSE}
devtools::install_github("RamiKrispin/coronavirus")
```
We can then include it and get updates with

```{r}
library(coronavirus)
update_dataset()
```

To use the dataset we can do the following:

```{r}
data("coronavirus")
```

And then take a look and the data we have available

```{r}
glimpse(coronavirus)
```

Now let's try a simple plot of UK confirmed case count

```{r cv-uk-case-count}
uk <- coronavirus %>%
  filter(country == "United Kingdom") %>%
  filter(type == "confirmed" & cases > 0 & province == "") %>%
  select(-lat, -long, -country) %>%
  arrange(date)

uk
```

```{r}
ggplot(uk, aes(date, cases)) +
  geom_line() +
  geom_point(alpha=0.2) +
  geom_smooth(se=FALSE) +
  labs(
    title = "Confirmed UK Coronavirus Cases",
    x = "Confirmed Cases",
    y = "Date"
  ) +
  theme_tufte()
```



Some simple examples

**Top 20 countries**

```{r}
library(dplyr)

summary_df <- coronavirus %>% 
  filter(type == "confirmed") %>%
  group_by(country) %>%
  summarise(total_cases = sum(cases)) %>%
  arrange(-total_cases)
summary_df
```

New cases in past 24 hours

```{r}
library(tidyr)
coronavirus %>% 
  filter(date == max(date)) %>%
  select(country, type, cases) %>%
  group_by(country, type) %>%
  summarise(total_cases = sum(cases)) %>%
  pivot_wider(names_from = type,
              values_from = total_cases) %>%
  arrange(-confirmed)
```

Plotting cases

```{r}
library(plotly)

coronavirus %>% 
  group_by(type, date) %>%
  summarise(total_cases = sum(cases)) %>%
  pivot_wider(names_from = type, values_from = total_cases) %>%
  arrange(date) %>%
  mutate(active = confirmed - death - recovered) %>%
  mutate(active_total = cumsum(active),
                recovered_total = cumsum(recovered),
                death_total = cumsum(death)) %>%
  plot_ly(x = ~ date,
                  y = ~ active_total,
                  name = 'Active', 
                  fillcolor = '#1f77b4',
                  type = 'scatter',
                  mode = 'none', 
                  stackgroup = 'one') %>%
  add_trace(y = ~ death_total, 
             name = "Death",
             fillcolor = '#E41317') %>%
  add_trace(y = ~recovered_total, 
            name = 'Recovered', 
            fillcolor = 'forestgreen') %>%
  layout(title = "Distribution of Covid19 Cases Worldwide",
         legend = list(x = 0.1, y = 0.9),
         yaxis = list(title = "Number of Cases"),
         xaxis = list(title = "Source: Johns Hopkins University Center for Systems Science and Engineering"))
```

**Teemap plot**

```{r}
conf_df <- coronavirus %>% 
  filter(type == "confirmed") %>%
  group_by(country) %>%
  summarise(total_cases = sum(cases)) %>%
  arrange(-total_cases) %>%
  mutate(parents = "Confirmed") %>%
  ungroup() 
  
  plot_ly(data = conf_df,
          type= "treemap",
          values = ~total_cases,
          labels= ~ country,
          parents=  ~parents,
          domain = list(column=0),
          name = "Confirmed",
          textinfo="label+value+percent parent")
```


## 3. Basic global daily trends

Let's create a date rather than having day, month, year

```{r}
data$Date <- as.Date(paste(data$year, data$month, data$day, sep="-"), "%Y-%m-%d")
glimpse(data)
```


Let's create a date and calculate the cumulative total for each day

```{r}
daily_global_counts <- data %>%
  group_by(Date) %>%
  summarise(
    total_cases = sum(cases),
    total_deaths = sum(deaths)
  )
daily_global_counts
```

We do a quick sanity check by plotting this data on a line graph (as it is time based)

```{r}
ggplot(daily_global_counts, aes(Date, total_cases)) +
  geom_point(alpha=0.4) +
  geom_line() +
  geom_smooth(se=FALSE, alpha=0.5, linetype=3) +
  labs(
    title = "Global COVID-19 Cases\nper day"
  ) +
  ylab('New Cases') + xlab('Date') +
  theme_minimal()
```

And the same for deaths

```{r}
ggplot(daily_global_counts, aes(Date, total_deaths)) +
  geom_point(alpha = 0.2) +
  geom_line() +
  geom_smooth(se=FALSE) +
  labs(
    title = "Cumulative COVID-19 Cases\nper day",
    ylab = "Deaths",
    xlab = "Date"
  ) +
  theme_minimal()
```


## Cumulative Global Data

The table above shows the cumulative confirmed cases of COVID-19 worldwide by date. Just reading numbers in a table makes it hard to get a sense of the scale and growth of the outbreak. Let's draw a line plot to visualize the confirmed cases worldwide.

```{r}
daily_global_counts
```

We do the cumulative counts

```{r}
daily_global_counts$cum_cases = cumsum(daily_global_counts$total_cases)
daily_global_counts$cum_deaths = cumsum(daily_global_counts$total_deaths)
```

And plot

```{r}
cum_cases_plot <- ggplot(daily_global_counts, aes(Date, cum_cases)) +
  geom_point() + 
  geom_line() +
  labs(xlab = "Date", ylab="Cumulative confirmed cases", title="Cumulative COVID-19 confirmed cases over time")

cum_cases_plot
```

And the log10 version on the y axis

```{r}
cum_cases_plot + 
  scale_y_log10() +
  labs(
    title = "Cumulative cases (log10)",
    caption = "Source: "
  )
```

## By country



Let's now group by countries and look at individual national trends across the world

```{r}
by_country <- data %>%
  group_by(countriesAndTerritories) %>%
  summarise(
    total_cases = sum(cases),
    total_deaths = sum(deaths)
  )
by_country

```

This is useful but it would also be useful to have some cumulative case and death counts for every country so that we can filter countries we wish to examine

```{r}
top_cases_by_country <- by_country %>%
  filter(total_cases > 100000) %>%
  arrange(desc(total_cases))
top_cases_by_country
```

We can show this on a vertical barchart

```{r}

library(RColorBrewer)
palette <- brewer.pal(5, "RdYlBu")[-(2:4)]

global_mean <- median(top_cases_by_country$total_cases)
x_start <- global_mean + 20000
y_start <- 5.5
x_end <- global_mean
y_end <- 7.5

ggplot(top_cases_by_country, aes(total_cases, countriesAndTerritories, color=total_cases)) +
  geom_point(size = 6) +
  geom_segment(aes(xend=100000, yend=countriesAndTerritories), size=1) +
  geom_text(aes(label = total_cases), color="white", size=1.5) +
  scale_x_continuous("", expand=c(0,0), limits=c(50000,2000000), position="top") +
  scale_color_gradientn(colors = palette) +
  labs(
    title = 'Top Countries by Confirmed Cases',
    captions = 'Source: opendata.org'
  ) +
  theme_minimal() +
  theme(
    axis.line.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title = element_blank(),
    axis.text = element_text(color = "black"),
    legend.position = "none"
  ) +
  geom_vline(xintercept = global_mean, color="grey40", linetype=3) +
    annotate(
    "text", 
    x = x_start, 
    y = y_start, 
    label="The\nglobal\naverage",
    vjust = 1, size = 3, color="grey40"
  ) +
  annotate(
    "curve",
    x = x_start, y=y_start,
    xend = x_end, yend = y_end,
    arrow = arrow(
      length = unit(0.2, "cm"),
      type = "closed"
    ),
    color="grey40"
  )
  
```


## 4. Normal vs under-reporting of deaths

Let's make a faceted view for each country of the cumulative cases
Displaying multiple countries as a facet


```{r}

by_country <- data %>%
  group_by(countriesAndTerritories, Date) %>%
  filter(countriesAndTerritories %in% c('China', 'Italy', 'Sweden', 'France', 'Iran', 'Brazil', 'India')) %>%
  summarise(
    total_cases = sum(cases),
    total_deaths = sum(deaths)
  ) %>%
  mutate(
    cum_cases = cumsum(total_cases),
    cum_deaths = cumsum(total_deaths)
  )
  
by_country

ggplot(by_country) + 
  geom_line(aes(Date, total_cases), color='blue') +
  facet_wrap(~countriesAndTerritories, ncol=2, scale="free_y") +
  geom_col(aes(Date, total_deaths), color='red') +
  labs(
    title="Cases and Deaths for COVID-19",
    subtitle="Shown for select countries with stark\ndifference between reported case/death ratios",
    ylab = "Cases & Deaths"
  ) +
  theme_clean()
  
  
  
  
  
```

It is clear that in some countries there under-reporting of deaths. Let's separate out some key suspects

```{r}
under_reporting <- by_country %>%
  filter(countriesAndTerritories %in% c('China', 'Brazil', 'Iran'))

ggplot(under_reporting) + 
  geom_line(aes(Date, total_cases), color="steelblue") +
  facet_wrap(~countriesAndTerritories) +
  geom_col(aes(Date, total_deaths), color='red') +
  labs(
    title="Countries potentially underreporting case/death ratio"
  ) +
  theme_bw()
```

Compared to more typical countries

```{r}
ord_reporting <- by_country %>%
  filter(countriesAndTerritories %in% c('France', 'Italy', 'United_Kingdom'))

ggplot(ord_reporting) + 
  geom_line(aes(Date, total_cases), color="steelblue") +
  facet_wrap(~countriesAndTerritories) +
  geom_col(aes(Date, total_deaths), color='red') +
  labs(
    title="Countries reporting typical case/death ratios"
  ) +
  theme_bw()
```

## 5. UK Cases Rate and Rolling Functions

Prepare the data as usual by getting 

```{r}
uk_data <- data %>%
  filter(countriesAndTerritories == 'United_Kingdom') %>%
  group_by(Date) %>%
  arrange(Date) %>%
  summarise(
    total_cases = sum(cases),
    total_deaths = sum(deaths)
  ) %>%
  mutate(
    cum_cases = cumsum(total_cases),
    cum_deaths = cumsum(total_deaths)
  )
  # summarize(
  #   cum_sizes  = cumsum(cases),
  #   cum_deaths = cumsum(deaths)
  # )
uk_data
```

Now let's plot the total daily cases and the cumulative cases:

```{r}
require(gridExtra)
plot1 <- ggplot(uk_data, aes(Date, total_cases)) +
  geom_line() +
  labs(
    title = 'UK Cases per day',
    subtitle = 'Jan, 2020 to June 2020',
    caption = 'Source: Open Data',
    ylab = 'Total cases'
  ) +
  ylim(0, 7500)
plot2 <- ggplot(uk_data, aes(Date, total_deaths)) +
  geom_line() +
  labs(
    title = 'UK Deaths per day',
    subtitle = 'Jan, 2020 to August 2020',
    caption = 'Source: Open Data',
    ylab = 'Total Deaths'
  ) +
  ylim(0, 7500)

grid.arrange(plot1, plot2, nrow=1)
```

```{r}
pacman::p_load(
  tidyquant,
  zoo
)

uk_roll_mean <- uk_data %>%
  tq_mutate(
    select = total_cases,
    mutate_fun = rollapply,
    width = 30,
    align = 'right',
    FUN = mean,
    na.rm = TRUE,
    col_rename = "mean_30"
  ) %>%
 tq_mutate(
    select     = total_cases,
    mutate_fun = rollapply,
    width      = 90,
    align      = "right",
    FUN        = mean,
    na.rm      = TRUE,
    col_rename = "mean_90"
  )

uk_roll_mean %>%
  ggplot(aes(Date, total_cases)) +
    # geom_point(alpha=0.2) +
    geom_line(aes(Date, mean_30), color= palette_light()[[1]], size = 1, linetype = 1) +
    geom_line(aes(Date, mean_90), color= palette_light()[[2]], size = 1, linetype = 1) +
    labs(
      title = "UK Cases per day",
      subtitle = "Showing 30 and 90 days moving averages"
    ) +
    scale_color_tq() +
    theme(legend.position = "none")
    

```

### Rolling custom functions

```{r}
# Custom function to return mean, sd, 95% conf interval
custom_rolling_stats <- function(x, na.rm = TRUE) {
    # x     = numeric vector
    # na.rm = boolean, whether or not to remove NA's
    
    m  <- mean(x, na.rm = na.rm)
    s  <- sd(x, na.rm = na.rm)
    hi <- m + 2*s
    lo <- m - 2*s
    
    ret <- c(mean = m, stdev = s, hi.95 = hi, lo.95 = lo) 
    return(ret)
}
```

Now we can use this function on our UK cases data

```{r}
uk_data_rollstats <- uk_data %>%
  tq_mutate(
    select = total_cases,
    mutate_fun = rollapply,
    width = 30,
    align = "right",
    by.column = FALSE,
    FUN = custom_rolling_stats,
    na.rm = TRUE
  )

uk_data_rollstats
```

We now have the data to view

1. Rolling average (trend)
2. 95% confidence interval (volatility)

Let's plot this in a Bollinger Bands style

```{r}
uk_data_rollstats %>%
  ggplot(aes(Date)) +
  geom_point(aes(y=total_cases), color="grey40", alpha=0.5) +
  geom_ribbon(aes(ymin = lo.95, ymax = hi.95), alpha=0.4) +
  geom_line(aes(y = mean), linetype = 2, size = 1, alpha = 0.7, color="red") +
  labs(
    title = "UK Cases by day", x = "",
    subtitle = "30-Day Moving Average with\n 95% Confidence Interval Bands (+/-2 Standard Deviations)"
  ) + 
  scale_color_tq() +
  theme_tq() +
  theme(
    legend.position = "None"
  )
  
```


Now let's try this and play with a variable for the width of the rolling average called ```roll_width()``` and also perform the analysis on death data


```{r}

rollwidth <- 7

# cases plot
plot_1 <- uk_data %>%
  tq_mutate(
    select = total_cases,
    mutate_fun = rollapply,
    width = rollwidth,
    align = "left",
    by.column = FALSE,
    FUN = custom_rolling_stats,
    na.rm = TRUE
  ) %>%
  ggplot(aes(x = Date)) +
  geom_point(aes(y=total_cases), color="grey40", alpha=0.5, position = 'stack') +
  geom_ribbon(aes(ymin = lo.95, ymax = hi.95), alpha=0.4) +
  geom_line(aes(y = mean), linetype = 1, size = 1, alpha = 0.8, color="steelblue") +
  labs(
    title = "UK Cases by day", x = "",
    subtitle = "Dynamic Moving Average with\n 95% Confidence Interval Bands (+/-2 Standard Deviations)",
    caption = "Open Data",
    ylab = "Total Cases"
  ) + 
  scale_color_tq() +
  theme_tq() +
  theme(
    legend.position = "None"
  )

# deaths plot
plot_2 <- uk_data %>%
  tq_mutate(
    select = total_deaths,
    mutate_fun = rollapply,
    width = rollwidth,
    align = "left",
    by.column = FALSE,
    FUN = custom_rolling_stats,
    na.rm = TRUE
  ) %>%
  ggplot(aes(x = Date)) +
  geom_point(aes(y=total_deaths), color="grey40", alpha=0.5, position = 'stack') +
  geom_ribbon(aes(ymin = lo.95, ymax = hi.95), alpha=0.4) +
  geom_line(aes(y = mean), linetype = 1, size = 1, alpha = 0.8, color="steelblue") +
  labs(
    title = "UK Deaths by day", x = "",
    subtitle = "Dynamic Moving Average with\n 95% Confidence Interval Bands (+/-2 Standard Deviations)",
    caption = "Open Data",
    ylab = "Total Cases"
  ) + 
  scale_color_tq() +
  theme_tq() +
  theme(
    legend.position = "None"
  )

grid.arrange(plot_1, plot_2, nrow=1)
```

Now let's try to normalise the x-axis and plot both on the same axis

```{r}

cases_roll_data <- uk_data %>%
  tq_mutate(
    select = total_cases,
    mutate_fun = rollapply,
    width = 7,
    align = "right",
    by.column = FALSE,
    FUN = custom_rolling_stats,
    na.rm = TRUE
  )

deaths_roll_data <- uk_data %>%
  tq_mutate(
    select = total_deaths,
    mutate_fun = rollapply,
    width = 7,
    align = "right",
    by.column = FALSE,
    FUN = custom_rolling_stats,
    na.rm = TRUE
  )

    
  ggplot(NULL, aes(x = Date)) +
  geom_point(data=cases_roll_data, aes(y=total_cases), color="grey40", alpha=0.5, position = 'stack') +
  geom_ribbon(data=cases_roll_data, aes(ymin = lo.95, ymax = hi.95), alpha=0.4) +
  geom_line(data=cases_roll_data, aes(y = mean), linetype = 1, size = 1, alpha = 0.8, color="steelblue") +
  geom_line(data=deaths_roll_data, aes(y = mean), linetype = 1, size = 0.5, alpha = 0.5, color="red") +
  geom_ribbon(data=deaths_roll_data, aes(ymin = lo.95, ymax = hi.95), alpha=0.4) +
  labs(
    title = "UK Cases & Deaths per day", x = "Date", y = "Cases & Deaths",
    subtitle = "Dynamic Moving Average with\n 95% Confidence Interval Bands (+/-2 Standard Deviations)",
    caption = "Open Data",
    ylab = "Total Cases"
  ) + 
  scale_color_tq() +
  theme_tq() +
  theme(
    legend.position = "None"
  )
```


```{r}
pacman::p_load(
  data.table
)
data_url <- "https://c19downloads.azureedge.net/downloads/csv/coronavirus-cases_latest.csv"
raw_data <- fread(data_url, check.names = TRUE)

plot_lab_confirmed_cases <- function(raw_data, area_name, area_type){
  area_data <- raw_data[
    Area.name == area_name &
      Area.type == area_type,,
    ][,Specimen.date := as.Date(Specimen.date)
      ][,c("Specimen.date","Daily.lab.confirmed.cases")][
        order(Specimen.date)
        ]
  area_data <- merge(area_data,
                       data.table(Specimen.date = seq(
                         min(area_data[,Specimen.date]),
                         max(area_data[,Specimen.date]),
                         by = "1 day"
                       )), all = TRUE, by = "Specimen.date")
  setkey(area_data, Specimen.date)
  setnafill(area_data, type = "const", fill = 0,
            cols = c("Daily.lab.confirmed.cases"))
  area_data[,roll_mean := frollmean(Daily.lab.confirmed.cases, n = 7, align = "right")]
  m_area_data <- melt(area_data, id.vars="Specimen.date",
                        measure.vars = c("Daily.lab.confirmed.cases","roll_mean"))
  area_plot <- ggplot(m_area_data, aes(x = Specimen.date, y = value, fill = variable, color = variable))+
    geom_bar(data = subset(m_area_data, variable == "Daily.lab.confirmed.cases"),
             stat = "identity") +
    geom_line(data = subset(m_area_data, variable == "roll_mean")) +
    labs(x="Specimen Date", y="Number of Confirmed Cases",
         fill = "", color = "") +
    scale_fill_manual(values = c("#ff0000","#000000"),
                      labels = c(sprintf("%s # Daily Confirmed cases",area_name),
                                 "7 day average")) +
    scale_color_manual(values = c("#ff0000","#000000"),
                       labels = c(sprintf("%s # Daily Confirmed cases",area_name),
                                  "7 day average")) +
    scale_x_date(date_breaks = "2 weeks", date_labels = "%Y-%m-%d") +
    theme_bw() %+replace% theme(legend.position = "top",
                                legend.justification = "left")
  area_plot
}


plot_lab_confirmed_cases(raw_data, "London", "region")
# ggsave(filename = "London_COVID.png", london_plot,
#        width = 10, height = 6)
```

The some columns on the data file has changed so I have updated the code from my last post.

It is worrying that the daily number of new cases has risen to the same level as when the lock down started. The speed of the increase remains quite slow however. This can be because but not limited to:

* The level of immunity in London is higher
* People are more careful (but, as I observe, more and more relaxed as days go by)
* Restricted mobility and work from home
* Better ventilation as the weather has been very warm and people here have no air conditioning.

It is therefore quite tempting to conclude the critical threshold for the second wave is going to be higher than roughly 150 people per day when the last lock-down started. I am going to venture a guess - 200-300 people per day potentially? If the trend continues, mindful that the latest dip is due to lag, then the level can be breached in 2-4 weeks time?


## Hillingdon Cases

```{r}
pacman::p_load(
  data.table,
  
)
data_url <- "https://c19downloads.azureedge.net/downloads/csv/coronavirus-cases_latest.csv"
raw_data <- data.table::fread(data_url, check.names = TRUE)
hillingdon_data <- raw_data[
  Area.name == "Hillingdon" &
    Area.type == "ltla",,
  ][,Specimen.date := as.Date(Specimen.date)
    ][,c("Specimen.date","Daily.lab.confirmed.cases")][
           order(Specimen.date)
           ]
hillingdon_data <- merge(hillingdon_data,
                     data.table(Specimen.date = seq(
                       min(hillingdon_data[,Specimen.date]),
                       max(hillingdon_data[,Specimen.date]),
                       by = "1 day"
                     )), all = TRUE, by = "Specimen.date")
setkey(hillingdon_data, Specimen.date)
setnafill(hillingdon_data, type = "const", fill = 0,
          cols = c("Daily.lab.confirmed.cases"))
hillingdon_data[,roll_mean := frollmean(Daily.lab.confirmed.cases, n = 7, align = "right")]
m_hillingdon_data <- melt(hillingdon_data, id.vars="Specimen.date",
                      measure.vars = c("Daily.lab.confirmed.cases","roll_mean"))
ggplot(m_hillingdon_data, aes(x = Specimen.date, y = value, fill = variable, color = variable))+
  geom_bar(data = subset(m_hillingdon_data, variable == "Daily.lab.confirmed.cases"),
           stat = "identity") +
  geom_line(data = subset(m_hillingdon_data, variable == "roll_mean")) +
  labs(x="Specimen Date", y="Number of Confirmed Cases",
      fill = "", color = "") +
  scale_fill_manual(values = c("#ff0000","#000000"),
                    labels = c("Hillingdon # Daily Confirmed cases",
                               "7 day average")) +
  scale_color_manual(values = c("#ff0000","#000000"),
                    labels = c("Hillingdon # Daily Confirmed cases",
                               "7 day average")) +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%Y-%m-%d") +
  theme_bw() %+replace% theme(legend.position = "top",
                              legend.justification = "left")
# ggsave(filename = "Hillingdon_COVID.png", merton_plot, width = 10, height = 6)
```


